---
title: "GSE74548"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
theme: cosmo
---

## Load packages

```{r,packages}

knitr::opts_chunk$set(dev = "png")

suppressPackageStartupMessages({
    library("missMethyl")
    library("GEOquery")
    library("limma")
    library("topconfects")
    library("minfi")
    library("IlluminaHumanMethylation450kmanifest")
    library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
    library("DMRcate")
    library("mitch")

    library("kableExtra")
    library("forestplot")
    library("RColorBrewer")
    library("plyr")
    library("R.utils")
    library("eulerr")
    library("gplots")
    library("reshape2")
    library("beeswarm")
  })
# Annotation
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Loading functions

These functions provide shortcuts to help with charts and other analysis. They will
eventually be shoved into another Rscript or package but can stay here for now.

```{r,load_functions}

source("https://raw.githubusercontent.com/markziemann/ART_methylation/master/meth_functions.R")
myranks<-function(x) {
  x$score <- sign(x$logFC)/log10(x$adj.P.Val)
  y <- x[,"score",drop=FALSE]
  y$rn <- x$Row.names
  return(y)
}

# heatmap for continuous data
heatmap_c<-function(dm,name,mx,n, groups) {
  my_palette <- colorRampPalette(c("blue", "white", "red"))(25)
  topgenes <-  rownames(head(dm[order(dm$P.Value),],n))
  ss <- mx[which(rownames(mx) %in% topgenes),]
  mycols <- colorRampPalette(c("white","yellow", "orange", "red","darkred"))(n = length(groups))
  colCols <- mycols[order(groups)]
  heatmap.2(ss,scale="row",margin=c(10, 10),cexRow=0.6,trace="none",cexCol=0.4,
  ColSideColors=colCols ,  col=my_palette, main=name)
}

# heatmap for continuous data - topconfects
make_heatmap2_c <- function(confects,name,mx,n, groups) {
  topgenes <-  head(confects$table$name,n)
  my_palette <- colorRampPalette(c("blue", "white", "red"))(25)
  ss <- mx[which(rownames(mx) %in% topgenes),]
  mycols <- colorRampPalette(c("white","yellow", "orange", "red","darkred"))(n = length(groups))
  colCols <- mycols[order(groups)]
  heatmap.2(ss,scale="row",margin=c(10, 10),cexRow=0.6,trace="none",cexCol=0.4,
  ColSideColors=colCols ,  col=my_palette, main=name)  
}


run_mitch_1d <- function(dma,name) {
  library("mitch")
  rank <- run_mitch_rank(dma)
  capture.output(
    res <- mitch_calc(x = rank,genesets = genesets, priority = "significance",resrows=20)
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)
 
  capture.output(
    mitch_plots(res,outfile=paste(name,".pdf",sep=""))
    , file = "/dev/null", append = FALSE,
    type = c("output", "message"), split = FALSE)
  return(res$enrichment_result)

  nsig=nrow(subset(placebo_vs_supplement_followup_sex_age_mitch,p.adjustANOVA<0.05))

  if (nsig>2) {
    mitch_report(res=res , outfile=paste(name,".html",sep="") , overwrite = TRUE)
  }
}


```

## Data import

```{r,load}

load("AccesionnumberGSE74548.Rdata")

```

## Enrichment analysis

We will be using the recently published package mitch to perform enrichment analysis, using
average promoter methylation change as an indicator of gene activity.
Enrichment will be tested with the mitch package.

dma1a (placebo_vs_supplement_followup(sex,age))

dma2ba (placebo_vs_supplement_Baseline(sex,age))

dma3a (hcy_levels(sex,age))

dma4a (folate_levels(sex,age))

dma5a (vitb12 levels(sex,age))

```{r,genesets, fig.width = 8 ,fig.height = 8}

library("mitch")
# gene sets
download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
unzip("ReactomePathways.gmt.zip",overwrite = TRUE)
genesets <- gmt_import("ReactomePathways.gmt")

```

One dimensional enrichment analysis with REACTOME gene sets using average promoter methylation t-statistic.

```{r,1danalysis, fig.width = 8 ,fig.height = 8}

placebo_vs_supplement_followup_sex_age_mitch <- run_mitch_1d(dma= dma1a, name="placebo_vs_supplement_followup(sex,age)_mitch")
head(placebo_vs_supplement_followup_sex_age_mitch,50) %>% kbl() %>% kable_paper("hover", full_width = F)

placebo_vs_supplement_Baseline_sex_age_mitch <- run_mitch_1d(dma= dma2a, name="placebo_vs_supplement_Baseline_(sex,age)_mitch")
head(placebo_vs_supplement_Baseline_sex_age_mitch,50) %>% kbl() %>% kable_paper("hover", full_width = F)

hcy_levels_sex_age_mitch <- run_mitch_1d(dma= dma3a, name="hcy_levels(sex,age)_mitch")
head(hcy_levels_sex_age_mitch,50) %>% kbl() %>% kable_paper("hover", full_width = F)

folate_levels_sex_age_mitch <- run_mitch_1d(dma= dma4a, name="folate_levels(sex,age)_mitch")
head(folate_levels_sex_age_mitch,50) %>% kbl() %>% kable_paper("hover", full_width = F)

vitb12_levels_sex_age_mitch <- run_mitch_1d(dma= dma5a, name="vitb12 levels(sex,age)_mitch")
head(vitb12_levels_sex_age_mitch,50) %>% kbl() %>% kable_paper("hover", full_width = F)

```

# Run multi-dimensional enrichment analysis

```{r,mitch, fig.width = 8 ,fig.height = 8}

xl <- list("psf"=dma1a,"psb"=dma2a,"hcy"=dma3a,"folate"=dma4a,"vitb12"= dma5a)
xxl <- lapply(X = xl,run_mitch_rank)  
xxxl <- lapply(xxl,function(xxl) { xxl$genenames <- rownames(xxl) ; xxl} )
xxll <- join_all(xxxl,by="genenames")
rownames(xxll) <- xxll$genenames
xxll$genenames=NULL
colnames(xxll) <- names(xl)
head(xxll) %>% kbl() %>% kable_paper("hover", full_width = F)
capture.output(
        res <- mitch_calc(xxll,genesets = genesets,priority = "significance"),
        file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)

head(res$enrichment_result,50) %>% kbl() %>% kable_paper("hover", full_width = F) 

unlink("multi_mitch.pdf")
     capture.output(
        mitch_plots(res,outfile="multi_mitch.pdf")
        , file = "/dev/null", append = FALSE,
        type = c("output", "message"), split = FALSE)



     
length(which(res$enrichment_result$p.adjustMANOVA<0.05))
hm<-as.matrix(res$enrichment_result[1:40,4:8])
rownames(hm)<-res$enrichment_result[1:40,1]
heatmap.2(hm,trace ="none",scale = "none",dendrogram = "none",margins = c(7,25),cexCol = 0.7)

```

## Session information

```{r,sessioninfo}

sessionInfo()

```

