---
title: "GSE74548: Gene Methylation Enrichment Analysis"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

##  Introduction

Previously, Mandhri and I analysed the B-PROOF 450K data, trying to understand
whether vitamin supplementation caused changes in gene methylation.
We used Limma and some basic analyses, which showed no specific probes with FDR<0.05,
nor any DMRs.

In this analysis we will use the principle of Gene Set Enrichment Analysis, applying it
to many probes belonging to genes.
If the probes are trending in concert, then we can make some judgement about the
enrichment of those probes.
The statistical test used is the Wilcox test, which can be applied to competitive 
or self contained test types.

```{r,libs}

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")
library("tictoc")

CORES= detectCores()

```

## Reactome gene sets

```{r,reactome}

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

```

## Functions

```{r,gmea}

# Gene methylation enrichment analysis
calc_sc <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a[2],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  gme_res <- mclapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    wtselfcont <- wilcox.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=wtselfcont$p.value)
  } , mc.cores=CORES )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}


gmea_volc <- function(res) {
  sig <- subset(res,`fdr(sc)` < 0.05)
  plot(res$median , -log10(res$`p-value(sc)`) ,
    xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
    pch=19, cex=0.5, col="gray",main="self contained test")
  grid()
  points(sig$median , -log10(sig$`p-value(sc)`) ,
    pch=19, cex=0.5, col="red")
}


gmea_barplot <- function(res) {
  par(mfrow=c(1,2))
  n=50
  gs <- head(rownames(res),50)
  tstats <- lapply(gs, function(g) {
    df[which(df$genes==g),"tvals"]
  })
  names(tstats) <- gs
  tstats <- tstats[order(unlist(lapply(tstats,median)))]
  boxplot(tstats,horizontal=TRUE,las=1,
    main="smallest p-val(selfcont)",cex.axis=0.6,
    xlab="t-statistic")
  grid()
  sig <- subset(res,`fdr(sc)` < 0.05)
  gs <- head(rownames(sig[order(-abs(sig$median)),]),n)
  if ( length(gs) >2 ) {
    tstats <- lapply(gs, function(g) {
      df[which(df$genes==g),"tvals"]
    })
    names(tstats) <- gs
    tstats <- tstats[order(unlist(lapply(tstats,median)))]
    boxplot(tstats,horizontal=TRUE,las=1,
      main="biggest effect size(median)",cex.axis=0.6,
      xlab="t-statistic")
    grid()
  } else {
    plot(1)
    mtext("too few significant genes found")
  }
  par(mfrow=c(1,1))
}


```
## DM1 - post trial supplement versus placebo

### Whole gene analysis

```{r,dm1}

dm <- read.table("dma1a.tsv")
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t,breaks=seq(from=-10,to=10,by=1))
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="gmeawg_dm1.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
mitch_report(mres,outfile="gmeawg_mitch_dm1.html",overwrite=TRUE)

```

### Promoter

```{r,dma1promoter}

dm <- read.table("dma1a.tsv")
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t,breaks=seq(from=-10,to=10,by=1))
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res ,file="gmeapr_dm1.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
mitch_report(mres,outfile="gmeapr_mitch_dm1.html",overwrite=TRUE)

```

## DM3 effect of homocysteine

### Whole gene analysis

```{r,dm3}

dm <- read.table("dma3a.tsv")
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t,breaks=seq(from=-10,to=10,by=1))
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="gmeawg_dm3.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
mitch_report(mres,outfile="gmeawg_mitch_dm3.html",overwrite=TRUE)

```

### Promoter

```{r,dma3promoter}

dm <- read.table("dma1a.tsv")
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t,breaks=seq(from=-10,to=10,by=1))
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res ,file="gmeapr_dm3.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
mitch_report(mres,outfile="gmeapr_mitch_dm3.html",overwrite=TRUE)

```

## Session Information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```
